---
title: "Metro Employment Index Derivation"
author: "Mike Silva"
date: "Monday, November 03, 2014"
output: html_document
---

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(reshape2)
library(ggplot2)
library(scales)
```

This document explains how the metro employment index is derived.  This index is base on the Local Area Unemployment Statistics (LAUS) published by the U.S. Bureau of Labor Statistics (BLS).

We begin by downloading the data:  

```{r}
if(!file.exists('~/data')){
  dir.create(file.path('~/data'))              
}
if(!file.exists('~/data/LAUS')){
  dir.create(file.path('~/data/LAUS'))              
}

if(!file.exists('~/data/LAUS/la.series')){
 download.file('http://download.bls.gov/pub/time.series/la/la.series', '~/data/LAUS/la.series')              
}

if(!file.exists('~/data/LAUS/la.data.0.CurrentU00-04')){
 download.file('http://download.bls.gov/pub/time.series/la/la.data.0.CurrentU00-04', '~/data/LAUS/la.data.0.CurrentU00-04')              
}

if(!file.exists('~/data/LAUS/la.data.0.CurrentU05-09')){
 download.file('http://download.bls.gov/pub/time.series/la/la.data.0.CurrentU05-09', '~/data/LAUS/la.data.0.CurrentU05-09')              
}

if(!file.exists('~/data/LAUS/la.data.0.CurrentU10-14')){
 download.file('http://download.bls.gov/pub/time.series/la/la.data.0.CurrentU10-14', '~/data/LAUS/la.data.0.CurrentU10-14')              
}
```

Next we determine which series to include.  We want MSA level data (area_type_code = B) on employment (measure_code = 5) that are not seasonally adjusted (seasonal = U):

```{r}
## Read in the LAUS series data
la.series <- read.delim('~/data/LAUS/la.series') %>%
  filter(area_type_code == "B", measure_code == 5, seasonal == "U")

## Clean up the MSA names
la.series$msa.name <- gsub("Employment: ", "", as.character(la.series$series_title))
la.series$msa.name <- gsub("[:punct:(]U[:punct:)]", "", la.series$msa.name)
la.series$msa.name <- gsub(" Metropolitan Statistical Area ", "", la.series$msa.name)
la.series$msa.name <- gsub(" Metropolitan NECTA ", "", la.series$msa.name)

# Exclude Puerto Rico
la.series <- subset(la.series, !grepl(", PR", la.series$msa.name))

## Select the data we need for the analysis
la.series <- select(la.series, msa.name, series_id) %>%
  arrange(msa.name)
```

We then pull in the LAUS data identified in the previous step into a time series:

```{r}
data <- read.delim('~/data/LAUS/la.data.0.CurrentU00-04') %>%
  filter(series_id %in% la.series$series_id, period != "M13")
data$value <- as.numeric(as.character(data$value))

temp <- read.delim('~/data/LAUS/la.data.0.CurrentU05-09') %>%
  filter(series_id %in% la.series$series_id, period != "M13")
temp$value <- as.numeric(as.character(temp$value))

## Append rows
data <- rbind(data, temp)

temp <- read.delim('~/data/LAUS/la.data.0.CurrentU10-14') %>%
  filter(series_id %in% la.series$series_id, period != "M13")
temp$value <- as.numeric(as.character(temp$value))

## Append rows
data <- rbind(data, temp)
```

We then identify the month and year the data represent:

```{r}
str.date <- paste0(substr(data$period, 2,3), "-01-", data$year)
data$date <- as.Date(str.date, format="%m-%d-%Y")
```

We then decompose the time series into seasonal, trend and irregular components using Loess keeping the trend:

````{r}
## Transform from long to wide data frame
molten <- select(data, series_id, date, value) %>%
  melt(id=c('series_id','date'), na.rm=TRUE)
sa.data.wide <- dcast(molten, date ~ series_id, value.var="value") %>%
  na.omit()

## Create the data frame that will hold the seasonally adjusted data
sa.data <- as.data.frame(matrix(ncol=3, nrow=0))
names(sa.data) <- c('series_id','date','value')

## Loop through each series and seasonally adjust
for(i in 2:ncol(sa.data.wide)){
  ## Seasonally adjust the data
  sa <- stl(ts(sa.data.wide[,i], start=c(2000,1), frequency = 12), "per")
  ## Build the temporary data frame to hold the seasonally adjusted data
  series_id <-  names(sa.data.wide)[i]
  temp.df <- data.frame(c(as.character(series_id)), sa.data.wide[,1], as.numeric(sa$time.series[,"trend"]))
  names(temp.df) <- names(sa.data)
  ## Append seasonally adjusted data
  sa.data <- rbind(sa.data, temp.df)
}

sa.data <- merge(sa.data, la.series)
```

Finally we create an index using the begining of the time series and the pre-great recession levels:

```{r}
## 2000 Employment Levels
y2k <- filter(sa.data, date == "2000-01-01") %>%
  select(series_id, value)
names(y2k) <- c('series_id', 'y2k.value')
sa.data <- merge(sa.data, y2k)
sa.data$y2k.emp.indx <- (sa.data$value / sa.data$y2k)*100

## Pre Great Recession Employment Levels
pre.gr <- filter(sa.data, date == "2007-11-01") %>%
  select(series_id, value)
names(pre.gr) <- c('series_id', 'pre.gr.value')
sa.data <- merge(sa.data, pre.gr)
sa.data$emp.indx <- (sa.data$value / sa.data$pre.gr)*100
```

Metro employment relative to the January 2000 level  

```{r, echo=FALSE}
ggplot(sa.data, aes(x=date, y=y2k.emp.indx, group = series_id)) + geom_line(color=alpha("#222222", .1)) + scale_x_date() + theme(axis.title.x = element_blank()) + ylab("Employment Index (Jan 2000 = 100)")
```

Metro employment relative to the pre-Great Recession level  

```{r, echo=FALSE}
ggplot(sa.data, aes(x=date, y=emp.indx, group = series_id)) + geom_line(color=alpha("#222222", .1)) + scale_x_date() + theme(axis.title.x = element_blank()) + ylab("Employment Index (Nov 2007 = 100)")
```

```{r, echo=FALSE}
rm(temp.df, sa.data.wide, i, series_id, sa, y2k, pre.gr, molten, temp, str.date)
```